---
title: "Runtime and Ratings: Are Longer Horror Movies Better (or Scarier)?"
output:
  pdf_document: default
  html_document: default
date: "2024-10-15"
---


**Name**                | **Student ID**  
----------------------- | ---------------  
Jeffrey Foster          | 10175104  
Michael Moon            | 30270188  
Gagandeep Thind         | 30254133  
Yifeng (Danny) Liang    | 30072968


## Project Purpose

  Being inspired by the month of October and the Halloween season, we decided to focus our project on a shared passion amongst our group, Horror Movies. Movies can be a great way to escape reality and immerse yourself, and give an outlet to relieve stress in day to day life. Since we are a group of busy students with limited hours in the day, we wanted to know if it’s worth it to watch a long movie or if saving time on a movie won’t affect the viewing experience. Because of this idea, we decided to investigate how movie run time relates to the perceived quality of the movie by viewers. We want to know if short movies are rated as highly as long movies, and decide if we need to set more time aside to watch.

  The population of our dataset is **all Horror Movies that have been listed on IMDb**. Also known as the International Movies Database, IMDb is a popular online movie database and user reviewing website. This website provides a repository for hundreds of thousands of movies, tv shows, and other forms of entertainment programs, with detailed information on hundreds of variables ranging from the individuals involved in the making of the content, aspects of the content itself, and one of our critical variables, user ratings (IMDB | Help, n.d.-b). IMDB claims that, as of September 2024, they have 693,133 individual Movies within their database, and 255,604 entertainment programs within the Horror genre (Press Room - IMDB, n.d.).  Using their advanced title search, we have found that the population of all horror movies on the IMDB platform is 42,758 (Advanced Search, n.d.). Therefore, the population we have derived our sample from are the 42,758 Horror Movies that are listed on IMDB.

  Since IMDB collects information on hundreds if not thousands of variables, we’ve decided to choose one of the most relevant qualities about the movie as our variable of interest, the **Average IMDb movie user rating**. IMDB offers users with an account the ability to rate a movie on an ordinal scale from 1 to 10, with 10 being high and 1 being low, indicating their vote on the quality of the movie (IMDB | Help, n.d.). These ratings are then aggregated and averaged, creating the Average IMDB movie user rating score for each individual movie. Therefore, while each rating is based on an ordinal scale, the average of these ordinal scales is, in practice, a continuous variable. This figure can inform potential movie watchers on the potential quality of a movie before choosing to watch the movie or not. This variable can be leveraged as a decision factor to an individual choosing which movie to watch.
  
  The factor that we will be leveraging within our parameter test with our variable of interest is the Runtime of each individual movie. Using this as our factor, the parameter that we are interested in testing is **whether there is a significant difference between the average IMDb ratings of long horror movies and short horror movies**. This will inform us if we are able to enjoy quality, short movies, or if passionate horror movie goers are required to sit through longer films.

## Data Collection

Initially, we looked at a variety of datasets on Kaggle (kaggle.com) to start our search. This included data on remote work effects on mental health outcomes, mobile device usage measures, fish lengths and weights, and many more datasets investigated. From this source, many of our datasets of interest were synthetic data from machine learning algorithms based on real-world data. These datasets met many of our necessary conditions, so we attempted some introductory tests on the data. After completing introductory EDA analysis, we chose not to proceed with these datasets for our project analysis. We found they were non-representative of real-world data, and would lead to inappropriate conclusions if used. Furthermore, many of these datasets contained issues of sample size, non-representative sampling, and datasets with significant quality concerns, providing us adequate reason to shift our focus.

Next, we looked into public polling data on election voting intentions in Canada. With a Federal Election expected to be held in the coming year, we decided it would be interesting to find data on the public perception of each federal political party in Canada. We intended to complete a proportion analysis on who would vote or not vote for a specific party if an election were held today. Most data collection on public polling is completed by independent polling firms such as Abacus data, Nanos Research, Léger, Angus Reid, and others. Unfortunately, these polling companies do not provide open access to their datasets from polling surveys. The lack of access to sample studies made it difficult to find real-world sample data on the topic. Additionally, we found it challenging to find polling information from previous elections. Because of the lack of access to studies, we pivoted away from this idea.

We concluded our search when we found the dataset **“IMDb Horror: Chilling Movie Dataset”** on Kaggle (IMDB Horror: Chilling Movie Dataset, 2023). This dataset was collected directly from the IMDb website, and looks at a sample of 837 movies categorized as horror. The sample is a collection of horror movies that have more than 25,000 reviews on the website, ensuring that a sufficient rating sample size has been reached. Some of the key variable columns in the dataset include release date, genre, director, runtime, rating, and gross profits. The data is in a .csv format and the gross profits column contains some NaN values. All other columns contain complete data and do not require further cleaning. The data is licensed under CC0, which is a public domain, and has no copyright. With a dataset of interest and with appropriate size, sampling, origin, and structure found, we decided to conclude our search.

## Statistical Analysis 

First, we loaded in our dataset. We downloaded in here all relevant libraries (`mosaic` and `ggplot2`) and looked over each variable using a head function:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
data <- read.csv("Horror Movies IMDb.csv")
library(mosaic)
library(ggplot2)
head(data, 5)
```

### Exploratory Data Analysis (EDA)

To start our analysis, we conducted an introductory EDA to gain a better understanding of the data. Our main variables of interest include Rating and Runtime. One key definition we hoped to reach based on our EDA was appropriate constant runtime values to define “long” and “short” horror movies.  

```{r}
favstats(data$Runtime)
```
```{r}
ggplot(data, aes(x = Rating)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Movie Ratings", x = "Rating", y = "Frequency") +
  theme_minimal()
boxplot(data$Runtime,main="Box Plot of Runtime",ylab="Runtime (mins)",
        col="skyblue",border = "black")
ggplot(data, aes(x = Runtime, y = Rating)) +
  geom_point(alpha = 0.5, color = "indianred") +
  labs(title = "Runtime vs. Ratings", x = "Runtime (min)", y = "Rating") +
  stat_smooth(method = "lm", formula = y ~ x, geom ="smooth") +
  theme_minimal()

```

Based off our initial EDA analysis, we gained insight on distributions of ratings and how to best define our test parameters. We found the distribution of movie ratings to be left skewed, and that ratings follow what appears be a relatively normal distribution. We chose to define a "long" movie as any movie that is found to be longer than or equal to Q3 and a "short" movie as any movie that is found to be shorter than or equal to Q1:

long: $\geq Q3$, short: $\leq Q1$

Further, we created a scatter plot to test if there is any visual relationship between runtime and ratings using a line of best fit. We see that there is a slightly positive relationship, which would lead us to believe that longer movies may have higher ratings than shorter movies. 

One final EDA visualization we completed was a side-by-side box plot of long and short horror movie ratings. We define long and short horror movie runtimes here that will be used in further analysis. We wanted to better visualize any differences between long and short horror movie ratings to gain a clearer picture before conducting statistical inferences and conclusion:

```{r}
longQ3=filter(data,Runtime>=108)
shortQ1=filter(data,Runtime<=91)
longQ3R=longQ3$Rating
shortQ1R=shortQ1$Rating
boxplot(longQ3R, shortQ1R, 
        main="Box Plot of Long and Short Runtime Ratings", 
        names=c("Long Runtime", "Short Runtime"), 
        ylab="Rating", 
        col=c("skyblue", "lightgreen"), 
        border="black")
```
By looking at the box plots, we can definitely see there is some level of difference in ratings for movies that have a long runtime relative to a short runtime. To test this, we conducted hypothesis testing analysis on the mean difference between long and short horror movies to see if there is a statistical difference at a 95% confidence level:

### Test 1: Hypothesis Testing
Our null hypothesis is that there is no difference between mean long horror movie ratings and mean short horror movie ratings.\
Our alternative hypothesis is that there is a difference between mean long horror movie ratings and mean short horror movie ratings.\

$H_0: \mu_{Long} - \mu_{Short} = 0$\
$H_A: \mu_{Long} - \mu_{Short} \neq 0$\

$\alpha = 0.05$

For our test statistic, we chose to use a t-test statistic and compare the two to come to a more accurate conclusion. In t-tests, one of the key assumptions that must be met is normality. In order to check for normality, we created a Normal Probability Plot using stat_qq as a function of qqplot.

```{r message=FALSE, warning=FALSE}
ggplot(data=longQ3,aes(sample=longQ3R))+
  stat_qq(col="blue")+stat_qqline(col="red")+
  ggtitle("Normal Probability Plot for Long Movies")
ggplot(data=shortQ1,aes(sample=shortQ1R))+
  stat_qq(col="blue")+stat_qqline(col="red")+
  ggtitle("Normal Probability Plot for Short Movies")
```

Since the produced plots of long horror movies and short horror movies follows roughly a straight line through the middle of the data points, these data can be determined to conform to normal probability models.\

Another important factor when conducting a t-test on two samples is the sample variance. We must find out if the samples have equal variance or not to conclude if we should complete an independent sample t-test or a Welch t-test.\
```{r}
favstats(longQ3R)
favstats(shortQ1R)
cat("Standard diviation of two populations are unequal, therefore, their variances are unequal.")
```
Because both ratings of long and short movies are considered to follow normal probability models, and the variances in the two populations' data sets are unequal, we will use Welch's t-test to compute our test statistic and p-value.\
```{r}
t.test(longQ3R,shortQ1R,alternative="two.sided",conf.level=0.95,val.equal=FALSE)
```
p-value: 2.197e-11 < 0.05 ($\alpha$)

At a significance level of 0.05, we reject the null hypothesis that there is no difference between long horror movie ratings and short horror movie ratings in favor of the alternative hypothesis that there is a difference between long horror movie ratings and short horror movie ratings.\


### Test 2: Confidence Interval Testing

The confidence interval test is useful to get the range of values within our confidence interval. To conduct this test, we will again define our null and alternative hypotheses.  We will test to see if long horror movies have a different mean rating than short horror movies:\

$H_0: \mu_{Long} - \mu_{Short} = 0$\
$H_A: \mu_{Long} - \mu_{Short} \neq 0$\

$\alpha = 0.05$

To evaluate our confidence interval, we will create a bootstrap distribution to find our bootstrap confidence interval.

With bootstrapping, we generated many mean differences based on the difference between resampled ratings of long and short horror movies. In our bootstrapping, 2000 repeats of the mean difference between long movie ratings and short movie ratings were simulated to compute the distribution of the mean difference. We then computed the 95% confidence interval based on the simulated distribution of mean difference and visualized our distribution.\
```{r message=FALSE, warning=FALSE}
B=do(2000)*(mean(resample(longQ3R))-mean(resample(shortQ1R)))
LL = quantile(B$result, 0.025)
UL = quantile(B$result, 0.975)
ggplot(B,aes(x=result))+
  geom_histogram(fill="skyblue",col="black")+
  ylab("Frequency")+
  xlab("Mean Difference")+
  ggtitle("Bootstrap Distribution")+
  geom_vline(xintercept = UL, color = "red")+ 
  geom_vline(xintercept = LL, color = "red")
interval=quantile(B$result,c(0.025,0.975))
interval
```

With 95% confidence, the true mean difference between long movie ratings and short movie ratings is captured in the interval of (0.4258151, 0.7766885). 

Since the confidence interval does not capture 0, at a 0.05 significance level, we can reject the null hypothesis that the mean rating of long horror movies is equal to the mean rating of short horror movies in favour of the alternative hypothesis that there is a difference between long horror movies ratings and short horror movies ratings."

## Conclusion

Through our completed analysis, we have found a statistically significant difference in IMDb movie ratings between short and long horror movies. Based on the results from our hypothesis test and bootstrapping confidence interval, the longest 25% of horror movies tend to receive higher average IMDb user ratings than the shortest 25% of horror movies. As a group of passionate horror fans, we now have the knowledge that informs us that, to have a better likelihood of enjoying a horror movie, a relatively longer movie choice may result in a better viewing experience.

In progressing through this project and identifying our key findings, our group learned a great deal about the practical implementation of hypothesis testing and confidence intervals in real-world datasets and real-world scenarios with little guidance. Additionally, we gained strong experience identifying datasets and utilizing decision criteria to accept or reject potential datasets. This project has improved our ability to progress through the stages of a data science project, conceptualize relevant questions and problems, and use our newfound skills to develop tangible insights and solutions.

## References

Chen, G. (2024). EDA numerical summaries PDF document. University of Calgary, Central Authentication Service. https://d2l.ucalgary.ca/d2l/le/content/618086/viewContent/6613182/View

Chen, G. (2024). Estimation: Point and interval estimation. University of Calgary, Central Authentication Service. https://d2l.ucalgary.ca/d2l/le/content/618086/viewContent/6613184/View

Chen, G. (2024). Introduction to data visualization. University of Calgary, Central Authentication Service. https://d2l.ucalgary.ca/d2l/le/content/618086/viewContent/6613181/View

Chen, G. (2024). Statistical tests of hypothesis. University of Calgary, Central Authentication Service. https://d2l.ucalgary.ca/d2l/le/content/618086/viewContent/6613187/View

Comprehensive R Archive Network (CRAN). (n.d.). CRAN: Package latexpdf. https://cran.r-project.org/web/packages/latexpdf/index.html

IMDb. (n.d.). Advanced search. https://www.imdb.com/search/title/?title_type=feature&genres=horror&sort=user_rating,desc

IMDb. (n.d.). Press Room - IMDb. https://www.imdb.com/pressroom/stats/

IMDb. (n.d.). What is IMDb? https://help.imdb.com/article/imdb/general-information/what-is-imdb/G836CY29Z4SGNMK5?ref_=helpart_nav_1#

IMDB Horror: Chilling Movie Dataset. (2023, June). Kaggle. https://www.kaggle.com/datasets/shreyanshverma27/imdb-horror-chilling-movie-dataset?resource=download

R Core Team. (2021). R: A language and environment for statistical computing. R Foundation for Statistical Computing. https://www.R-project.org/

Wickham, H. (2016). ggplot2: Elegant graphics for data analysis. Springer-Verlag New York. https://ggplot2.tidyverse.org

Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen, T., Miller, E., Bache, S., Müller, K., Ooms, J., Robinson, D., Seidel, D., Spinu, V., Takahashi, K., Vaughan, D., Wilke, C., Woo, K., & Yutani, H. (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686. https://doi.org/10.21105/joss.01686

Wickham, H., François, R., Henry, L., Müller, K., & Vaughan, D. (2023). dplyr: A grammar of data manipulation (Version 1.1.4). R package. https://dplyr.tidyverse.org

Xie, Y. (2021). knitr: A general-purpose package for dynamic report generation in R. https://yihui.org/knitr/


